---
title: "assignment week 5"
output: html_document
date: "2025-03-21"
---

```{r}
suppressPackageStartupMessages({
  library(AnnotationHub)
  library(ensembldb)
  library(GenomicRanges)
  library(epiwraps)
  library(rtracklayer)
  library(ggplot2)
})

if (!requireNamespace("R.utils", quietly = TRUE)) {
  install.packages("R.utils")
}
library(R.utils)
library(ggplot2)

```

## Identifying the bivalent domains (H3K27me3 and H3K4me3 overlaps) from mouse embryonic stem cells (ESCs)

1) Download peaks for H3K27me3 and H3K4me3 in embryonic mouse stem cell line Bruce 4 and convert to GRanges object

```{r}
download.file("https://www.encodeproject.org/files/ENCFF105NKG/@@download/ENCFF105NKG.bed.gz", "H3K27me3_peaks.bed.gz")
download.file("https://www.encodeproject.org/files/ENCFF247GVM/@@download/ENCFF247GVM.bed.gz", "H3K4me3_peaks.bed.gz")

gunzip("H3K4me3_peaks.bed.gz")
gunzip("H3K27me3_peaks.bed.gz")

# Convert BED-file to GRanges object
bed_file <- read.table("H3K4me3_peaks.bed", header = FALSE, nrows = 10)
print(bed_file)
```

```{r}
# Keep only the first three columns (chromosome, start, end) and assign appropriate names
H3K4me3_peaks <- read.table("H3K4me3_peaks.bed", header = FALSE)[, 1:3]
colnames(H3K4me3_peaks) <- c("chr", "start", "end")

# Convert to GRanges object
H3K4me3_peaks_gr <- GRanges(
  seqnames = H3K4me3_peaks$chr,
  ranges = IRanges(start = H3K4me3_peaks$start, end = H3K4me3_peaks$end)
)

# Inspect the GRanges object
H3K4me3_peaks_gr
```

```{r}
# Repeat for H3K27me3

#Look at the BED-file
bed_file <- read.table("H3K27me3_peaks.bed", header = FALSE, nrows = 10)
print(bed_file)

```

```{r}
# Read the BED file and keep only the first three columns (chromosome, start, end) and assign appropriate names
H3K27me3_peaks <- read.table("H3K27me3_peaks.bed", header = FALSE)[, 1:3]
colnames(H3K27me3_peaks) <- c("chr", "start", "end")

# Convert to GRanges object
H3K27me3_peaks_gr <- GRanges(
  seqnames = H3K27me3_peaks$chr,
  ranges = IRanges(start = H3K27me3_peaks$start, end = H3K27me3_peaks$end)
)# Inspect the GRanges object
H3K27me3_peaks_gr
```
2) Overlap H3K27me3 and H3K4me3

```{r}
hist(width(H3K4me3_peaks_gr), breaks=75)
```
```{r}
hist(width(H3K27me3_peaks_gr), breaks=75)
```
```{r}
# Extract the overlapping regions
bivalents <- findOverlaps(H3K4me3_peaks_gr, H3K27me3_peaks_gr, type="any")

bivalent_regions <- pintersect(H3K4me3_peaks_gr[from(bivalents)],H3K27me3_peaks_gr[to(bivalents)] )
bivalent_regions # regions only where both marks are present


```

## What happens to these regions upon differentiation?

1) Downloading the H3K4me3 and H3K27me3 peaks from CH12.LX (B-cells) in mus musculus and converting to GRanges objects

```{r}
download.file("https://www.encodeproject.org/files/ENCFF786KFS/@@download/ENCFF786KFS.bed.gz", "H3K27me3_Bcell_peaks.bed.gz")
download.file("https://www.encodeproject.org/files/ENCFF496DOC/@@download/ENCFF496DOC.bed.gz", "H3K4me3_Bcell_peaks.bed.gz")

gunzip("H3K4me3_Bcell_peaks.bed.gz")
gunzip("H3K27me3_Bcell_peaks.bed.gz")

# Read the BED file and keep only the first three columns (chromosome, start, end) and assign appropriate names
H3K27me3_Bcell_peaks <- read.table("H3K27me3_Bcell_peaks.bed", header = FALSE)[, 1:3]
colnames(H3K27me3_Bcell_peaks) <- c("chr", "start", "end")
H3K4me3_Bcell_peaks <- read.table("H3K4me3_Bcell_peaks.bed", header = FALSE)[, 1:3]
colnames(H3K4me3_Bcell_peaks) <- c("chr", "start", "end")

# Convert to GRanges object
H3K27me3_Bcell_peaks_gr <- GRanges(
  seqnames = H3K27me3_Bcell_peaks$chr,
  ranges = IRanges(start = H3K27me3_Bcell_peaks$start, end = H3K27me3_Bcell_peaks$end))
H3K4me3_Bcell_peaks_gr <- GRanges(
  seqnames = H3K4me3_Bcell_peaks$chr,
  ranges = IRanges(start = H3K4me3_Bcell_peaks$start, end = H3K4me3_Bcell_peaks$end))
# Inspect the GRanges object
H3K27me3_Bcell_peaks_gr
H3K4me3_Bcell_peaks_gr
```

2) Overlapping the bivalent regions of the mESCs with H3K4me3, H3K27me3 and with both in Bcells

```{r}
# Overlapping bivalent regions in embryonic cells with H3K4me3 peaks in differentiated B cells
bivalents_H3K4me3 <- findOverlaps(bivalent_regions, H3K4me3_Bcell_peaks_gr, type="any")
bivalents_H3K4me3 #number of bivalent regions overlapping H3K4me3
```
648 bivalent regions in the mESCs overlap with H3K4me3 domains in the Bcells

```{r}
# Overlapping bivalent regions in embryonic cells with H3K27me3 peaks in differentiated B cells
bivalents_H3K27me3 <- findOverlaps(bivalent_regions, H3K27me3_Bcell_peaks_gr, type="any")
bivalents_H3K27me3 #number of bivalent regions overlapping H3K27me3
```
984 bivalent regions in the mESCs overlap with  H3K27me3 domains in the Bcells

```{r}
# Overlapping bivalent regions in embryonic cells with both histone marks in differentiated B cells
H3K4me3_H3K27me3 <- findOverlaps(H3K4me3_Bcell_peaks_gr, H3K27me3_Bcell_peaks_gr) # overlap H3K4me3 and H3K27me3 in Bcells
H3K4me3_H3K27me3_Bcells <-pintersect(H3K4me3_Bcell_peaks_gr[from(H3K4me3_H3K27me3)],H3K27me3_Bcell_peaks_gr[to(H3K4me3_H3K27me3)] ) # select only regions where both marks are present
H3K4me3_H3K27me3_Bcells

# overlap bivalent regions in mESCs with regions were both marks are present
bivalents_differentiated <- findOverlaps(bivalent_regions,H3K4me3_H3K27me3_Bcells , type="any")
bivalents_differentiated
```
176 bivalent regions in the mESCs overlap with H3K4me3 and H3K27me3 domains in the Bcells


## How many of the mESC bivalent domains are, in the differentiated cell type, overlapping either mark or their combination?

- 648 bivalent regions in the mESCs overlap with H3K4me3 domains in the Bcells
- 984 bivalent regions in the mESCs overlap with  H3K27me3 domains in the Bcells
- 176 bivalent regions in the mESCs overlap with H3K4me3 and H3K27me3 domains in the Bcells

