---
title: "Assignment week 12"
output: html_document
date: "2025-05-28"
---

1) Preparaton: loading packages and downloading peaks and interactions

```{r, prep}
library(GenomicRanges)
library(rtracklayer)
library(org.Hs.eg.db)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(ggplot2)

# p300 peaks from ENCODE
download.file("https://www.encodeproject.org/files/ENCFF165QGE/@@download/ENCFF165QGE.bed.gz", "p300.bed.gz")
peaks <- import("p300.bed.gz", format = "narrowPeak")
seqlevelsStyle(peaks) <- "UCSC"

# Load Salviato distal enhancer–target predictions
salviato_gr <- readRDS(url("https://ethz-ins.org/content/hg38.SalviatoDistalEnhancerTargets.GR.rds"))
seqlevelsStyle(salviato_gr) <- "UCSC"
```
2) Overlapping the p300 peaks with the distal enhancers

```{r, overlaps}
ov <- findOverlaps(peaks, salviato_gr)
peaks_with_targets <- peaks[queryHits(ov)]
targets <- salviato_gr[subjectHits(ov)]

# Attach predicted gene symbol(s)
mcols(peaks_with_targets)$target_symbol <- targets$target
```

3) Obtaining the TSS positions

```{r, positions}
# Get TSSs from TxDb
tss_all <- promoters(genes(TxDb.Hsapiens.UCSC.hg38.knownGene), upstream = 0, downstream = 1)

# Convert Entrez IDs → gene symbols
tss_all$gene_symbol <- mapIds(
  org.Hs.eg.db,
  keys = names(tss_all),
  column = "SYMBOL",
  keytype = "ENTREZID"
)
```
4) Matching predicted target symbols to TSSs

```{r, matching}
flat_peaks <- rep(peaks_with_targets, lengths(peaks_with_targets$target_symbol))
flat_peaks$target_symbol <- unlist(peaks_with_targets$target_symbol)

# Match gene symbol to TSS
tss_match_idx <- match(flat_peaks$target_symbol, tss_all$gene_symbol)
valid <- !is.na(tss_match_idx)
flat_peaks <- flat_peaks[valid]
matched_tss <- tss_all[tss_match_idx[valid]]
```

5) Determine if the peaks are upstream or downstream of the TSS

```{r, stream}
midpoint <- start(flat_peaks) + width(flat_peaks) / 2
tss_pos <- start(matched_tss)
strand_info <- as.character(strand(matched_tss))

is_upstream <- ifelse(strand_info == "+", midpoint < tss_pos, midpoint > tss_pos)

upstream_peaks <- flat_peaks[is_upstream]
downstream_peaks <- flat_peaks[!is_upstream]
```

6) Comparing the target gene to the nearest gene by distance

```{r, compare}
# Find nearest gene (symbol) to each peak
nearest_idx <- nearest(flat_peaks, tss_all)
nearest_symbols <- tss_all$gene_symbol[nearest_idx]

predicted_symbols <- flat_peaks$target_symbol

# Valid comparisons
valid2 <- !is.na(predicted_symbols) & !is.na(nearest_symbols)

# Calculate proportions
prop_upstream <- mean(predicted_symbols[valid2 & is_upstream] == nearest_symbols[valid2 & is_upstream])
prop_downstream <- mean(predicted_symbols[valid2 & !is_upstream] == nearest_symbols[valid2 & !is_upstream])

prop_upstream
prop_downstream
```






## In what proportion of the cases the predicted target is the closest gene?

- Among the upstream genes: 0.04659794 (ca. 4.9%)
- Among the downstream genes: 0.05375925 (ca. 5.3%)

## Plot a distribution of the distances between the peak and the (TSS of the) predicted target gene

7) plotting the distance distribution to target TSS

```{r, plotting}
distances <- abs(midpoint - tss_pos)
df <- data.frame(
  distance = distances,
  direction = ifelse(is_upstream, "Upstream", "Downstream")
)

ggplot(df, aes(x = distance, fill = direction)) +
  geom_histogram(bins = 50, alpha = 0.7, position = "identity") +
  scale_x_log10() +
  theme_minimal() +
  labs(title = "Distance to predicted target gene TSS", x = "Distance (bp, log scale)", y = "Number of peaks")
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
